Lattices of acyclic pipe dreams

### 2023-05-10 acyclicFacets/draft/nuTamari.tex
Added a little detail in proof of 4.19

The \defn{Rothe diagram} of a permutation~$\omega$ is the set $\set{(\omega(j),i)}{i<j,\ \omega(i)>\omega(j)}$ in matrix notation.
A permutation $\omega$ is called \defn{dominant} if its Rothe diagram is the Ferrer's diagram of a partition located at the top left corner $(1,1)$. Equivalently, a permutation~$\omega$ is dominant if and only if it is 132-avoiding. 
For a permutation~$\omega = \omega_1 \dots \omega_n \in \fS_n$, we denote by~$0\omega$ the permutation~$0 \omega_1 \dots \omega_n$ of~$\{0, 1, \dots, n\}$, and we consider here pipe dreams with pipes indexed by~$\{0, 1, \dots, n\}$. The Theorem~\ref{prob:nuAcyclicProperty} below states that all pipe dreams with permutation $0 \omega$ are acyclic if and only if $\omega$ is a dominant permutation.

@@

\begin{theorem}%[The acyclic property for $\nu$-Tamari lattices]
\label{prob:nuAcyclicProperty}
All pipe dreams with permutation $0 \omega$ are acyclic if and only if $\omega$ is a dominant permutation.
\end{theorem}

\begin{proof} For the forward direction, assume that $\omega$ is not dominant. 
Let $P_0$ be the greedy pipe dream of $0\omega$. It has no crossing in row 0, and cannot be a partition starting at row 1.
There must be a row $r>1$ with more crossing than in row $r-1$. 
Considering the rightmost column $c$ of $P$ with a crossing in row $r$,  the pipes $r<k$ are crossing at $(c,r)$, $\omega^{-1}(k)<\omega^{-1}(r)$ and are in contact in position $(c+1,r-1)$.
Let $Q$ be the pipe dream obtained from $P_0$ by the flip of the crossing at $(c,r)$ with the contact at $(c+1,r-1)$.
The contact of $Q$ at $(c,r)$ would gives $r\contactLess{Q} k$. In row $0$ we only have contacts and $\omega^{-1}(k)<\omega^{-1}(r)$, combining this  with Lemma~\ref{lem:rectangle} would give  that 
$k \contactLess{Q} r$, a contradiction, hence $Q$ is not acyclic.

For the reverse direction, assume we have a pipe dream $P$ that is not acyclic. The contact graph must contain a cycle $i_1 \to i_2\to\cdots\to i_m\to i_1$.
This is only possible if some pipe $i_j$ and $i_\ell$ with some contact on both side of the crossing.
Hence that crossing is not horizontally nor vertically connected, therefore Lemma~\ref{lem:hv_connected} gives that $\omega$ is not dominant.
\end{proof}

>>>>

The \defn{Rothe diagram} of a permutation~$\omega$ is the set $\set{(\omega(j),i)}{i<j,\ \omega(i)>\omega(j)}$ in matrix notation.
A permutation $\omega$ is called \defn{dominant} if its Rothe diagram is the Ferrer's diagram of a partition located at the top left corner $(1,1)$. Equivalently, a permutation~$\omega$ is dominant if and only if it is 132-avoiding. 
For a permutation~$\omega = \omega_1 \dots \omega_n \in \fS_n$, we denote by~$0\omega$ the permutation~$0 \omega_1 \dots \omega_n$ of~$\{0, 1, \dots, n\}$, and we consider here pipe dreams with pipes indexed by~$\{0, 1, \dots, n\}$. The Theorem~\ref{prob:nuAcyclicProperty} below states that all pipe dreams with permutation $0 \omega$ are acyclic if and only if $\omega$ is a dominant permutation.
\nantel{modify intro to reflect this?}

@@

\begin{theorem}%[The acyclic property for $\nu$-Tamari lattices]
\label{prob:nuAcyclicProperty}
All pipe dreams with permutation $0 \omega$ are acyclic if and only if $\omega$ is a dominant permutation.
\end{theorem}
\nantel{add in acknowledgement that : Lucas Gagnon had a proof of converse in different contaxt}

\begin{proof} For the forward implication, assume that $\omega$ is not dominant. 
Let $P_0$ be the greedy pipe dream of $0\omega$. It has no crossing in row 0, and cannot be a partition starting at row 1.
There must be a row $r>1$ with more crossing than in row $r-1$. 
Considering the rightmost column $c$ of $P$ with a crossing in row $r$,  the pipes $r<k$ are crossing at $(c,r)$, $\omega^{-1}(k)<\omega^{-1}(r)$ and are in contact in position $(c+1,r-1)$.
Let $Q$ be the pipe dream obtained from $P_0$ by the flip of the crossing at $(c,r)$ with the contact at $(c+1,r-1)$.
The contact of $Q$ at $(c,r)$ would gives $r\contactLess{Q} k$. In row $0$ we only have contacts and $\omega^{-1}(k)<\omega^{-1}(r)$, combining this  with Lemma~\ref{lem:rectangle} would give  that 
$k \contactLess{Q} r$, a contradiction, hence $Q$ is not acyclic.

For the converse, assume we have a pipe dream that is not acyclic. The contact graph must contain a cycle $i_1 \to i_2\to\cdots\to i_m\to i_1$.
The contact $i_1 \to i_2$ implies that a portion of the pipe $i_2$ is southeast of $i_1$. The same is true for the contact $i_2\to i_3$ and so on.
In the end, we have that $i_m\to i_1$ gives a portion of the pipe $i_m$ is northwest of $i_1$. 
This is only possible if some pipe $i_j$ cross $i_1$ with the contact $i_1\to i_2$ on one side of the crossing of $i_1$ and $i_j$, and the contact $i_j \to i_{j+1 \mod m}$ on the other side.
Hence that crossing is not horizontally nor vertically connected, therefore Lemma~\ref{lem:hv_connected} gives that $\omega$ is not dominant.
\end{proof}
